---
title: "DGP Parameter Signal Strength Analysis"
author: "Analyst"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(here)
library(survival)
library(glmnet)
library(knitr)
library(kableExtra)

source(here("R/data-handler.R"))
source(here("R/datagen-helper.R"))
source(here("scripts/TV-CSL/time-varying-estimate.R"))
```

## Overview

This analysis investigates the signal strength of DGP parameters using different modeling approaches:
1. Lasso regression with different lambda values
2. Ridge regression
3. Standard Cox regression with varying sample sizes (n=500 and n=2000)

## Data Preparation

```{r data-prep}
intercept <- -1.5
slope_multiplier <- -0.5
true_coefficients <- slope_multiplier * 1/(1:10)
seed_value <- 42

file_path <- here("scripts", "datagen", "parameter-tuning", "data",
                 "eta-type-10-dim-linear_CATE-type-zero_n-500.rds")

dir.create(dirname(file_path), recursive = TRUE, showWarnings = FALSE)

if (!file.exists(file_path)) {
  train_data <- generate_simulated_data(
    n = 500,
    lambda_C = 0.1,
    eta_type = "10-dim-linear",
    CATE_type = "zero",
    seed_value = seed_value,
    linear_intercept = intercept,
    linear_slope_multiplier = slope_multiplier,
    verbose = 0
  )
  
  train_data_pseudo <- preprocess_data(
    single_data = train_data,
    run_time_varying = TRUE
  )
  
  saveRDS(
    list(
      train_data = train_data,
      train_data_pseudo = train_data_pseudo
    ),
    file = file_path
  )
} else {
  loaded_data <- readRDS(file_path)
  train_data <- loaded_data$train_data
  train_data_pseudo <- loaded_data$train_data_pseudo
}
```

## Model Fitting and Evaluation

```{r model-fitting}
regressor <- transform_X(
  single_data = train_data_pseudo,
  regressor_spec = "linear-only"
)

m_lasso_default <- cv.glmnet(
  regressor,
  Surv(train_data_pseudo$tstart, train_data_pseudo$tstop, train_data_pseudo$Delta),
  alpha = 1,
  family = "cox"
)

plot(m_lasso_default, main = "Lasso: Default Lambda Path")

m_beta_default <- coef(m_lasso_default, s = "lambda.min")
rmse_lasso_default <- sqrt(mean((true_coefficients - as.vector(m_beta_default))^2))

m_lasso_custom <- cv.glmnet(
  regressor,
  Surv(train_data_pseudo$tstart, train_data_pseudo$tstop, train_data_pseudo$Delta),
  alpha = 1,
  family = "cox",
  lambda = 10^seq(3, -2, by = -0.1)
)

plot(m_lasso_custom, main = "Lasso: Custom Lambda Path")

m_beta_custom <- coef(m_lasso_custom, s = "lambda.min")
rmse_lasso_custom <- sqrt(mean((true_coefficients - as.vector(m_beta_custom))^2))

m_ridge <- cv.glmnet(
  regressor,
  Surv(train_data_pseudo$tstart, train_data_pseudo$tstop, train_data_pseudo$Delta),
  alpha = 0,
  family = "cox",
  lambda = 10^seq(3, -2, by = -0.1)
)

plot(m_ridge, main = "Ridge Regression Lambda Path")

m_beta_ridge <- coef(m_ridge, s = "lambda.min")
rmse_ridge <- sqrt(mean((true_coefficients - as.vector(m_beta_ridge))^2))

m_cox <- coxph(
  Surv(tstart, tstop, Delta) ~ X.1 + X.2 + X.3 + X.4 + X.5 + X.6 + X.7 + X.8 + X.9 + X.10,
  data = train_data_pseudo
)

m_beta_cox <- coef(m_cox)
rmse_cox_n_500 <- sqrt(mean((true_coefficients - as.vector(m_beta_cox))^2))
```

```{r larger-sample}
file_path_n_2000 <- here("scripts", "datagen", "parameter-tuning", "data",
                        "eta-type-10-dim-linear_CATE-type-zero_n-2000.rds")

if (!file.exists(file_path_n_2000)) {
  train_data_2000 <- generate_simulated_data(
    n = 2000,
    lambda_C = 0.1,
    eta_type = "10-dim-linear",
    CATE_type = "zero",
    seed_value = seed_value,
    linear_intercept = intercept,
    linear_slope_multiplier = slope_multiplier,
    verbose = 0
  )
  
  train_data_pseudo_2000 <- preprocess_data(
    single_data = train_data_2000,
    run_time_varying = TRUE
  )
  
  saveRDS(
    list(
      train_data = train_data_2000,
      train_data_pseudo = train_data_pseudo_2000
    ),
    file = file_path_n_2000
  )
} else {
  loaded_data_2000 <- readRDS(file_path_n_2000)
  train_data_pseudo_2000 <- loaded_data_2000$train_data_pseudo
}

m_cox_n_2000 <- coxph(
  Surv(tstart, tstop, Delta) ~ X.1 + X.2 + X.3 + X.4 + X.5 + X.6 + X.7 + X.8 + X.9 + X.10,
  data = train_data_pseudo_2000
)

m_beta_cox_n_2000 <- coef(m_cox_n_2000)
rmse_cox_n_2000 <- sqrt(mean((true_coefficients - as.vector(m_beta_cox_n_2000))^2))
```
```{r coefficient-output}
print_coef_comparison <- function(true_coef, fitted_coef, model_name) {
  comparison_df <- data.frame(
    True = signif(true_coef, 3),
    Fitted = signif(as.vector(fitted_coef), 3)
  )
  cat("\n", model_name, "\n")
  print(kable(comparison_df, col.names = c("True Coefficients", "Fitted Coefficients")))
}

print_coef_comparison(true_coefficients, m_beta_default, "Lasso Default Lambda")
print_coef_comparison(true_coefficients, m_beta_custom, "Lasso Custom Lambda")
print_coef_comparison(true_coefficients, m_beta_ridge, "Ridge Regression")
print_coef_comparison(true_coefficients, m_beta_cox, "Cox Regression (n=500)")
print_coef_comparison(true_coefficients, m_beta_cox_n_2000, "Cox Regression (n=2000)")
```

```{r stronger-signal}
intercept_strong <- 3
slope_multiplier_strong <- 2.5
true_coefficients_strong <- slope_multiplier_strong * 1/(1:10)

file_path_strong <- here("scripts", "datagen", "parameter-tuning", "data",
                        "eta-type-10-dim-linear_CATE-type-zero_n-500_strong.rds")
file_path_n_2000_strong <- here("scripts", "datagen", "parameter-tuning", "data",
                               "eta-type-10-dim-linear_CATE-type-zero_n-2000_strong.rds")

# Generate n=500 data with stronger signal
if (!file.exists(file_path_strong)) {
  train_data_strong <- generate_simulated_data(
    n = 500,
    lambda_C = 0.1,
    eta_type = "10-dim-linear",
    CATE_type = "zero",
    seed_value = seed_value,
    linear_intercept = intercept_strong,
    linear_slope_multiplier = slope_multiplier_strong,
    verbose = 0
  )
  
  train_data_pseudo_strong <- preprocess_data(
    single_data = train_data_strong,
    run_time_varying = TRUE
  )
  
  saveRDS(
    list(
      train_data = train_data_strong,
      train_data_pseudo = train_data_pseudo_strong
    ),
    file = file_path_strong
  )
} else {
  loaded_data_strong <- readRDS(file_path_strong)
  train_data_pseudo_strong <- loaded_data_strong$train_data_pseudo
}

regressor_strong <- transform_X(
  single_data = train_data_pseudo_strong,
  regressor_spec = "linear-only"
)

# Fit models with stronger signal
m_lasso_default_strong <- cv.glmnet(
  regressor_strong,
  Surv(train_data_pseudo_strong$tstart, train_data_pseudo_strong$tstop, train_data_pseudo_strong$Delta),
  alpha = 1,
  family = "cox"
)

m_beta_default_strong <- coef(m_lasso_default_strong, s = "lambda.min")
rmse_lasso_default_strong <- sqrt(mean((true_coefficients_strong - as.vector(m_beta_default_strong))^2))

m_lasso_custom_strong <- cv.glmnet(
  regressor_strong,
  Surv(train_data_pseudo_strong$tstart, train_data_pseudo_strong$tstop, train_data_pseudo_strong$Delta),
  alpha = 1,
  family = "cox",
  lambda = 10^seq(3, -2, by = -0.1)
)

m_beta_custom_strong <- coef(m_lasso_custom_strong, s = "lambda.min")
rmse_lasso_custom_strong <- sqrt(mean((true_coefficients_strong - as.vector(m_beta_custom_strong))^2))

m_ridge_strong <- cv.glmnet(
  regressor_strong,
  Surv(train_data_pseudo_strong$tstart, train_data_pseudo_strong$tstop, train_data_pseudo_strong$Delta),
  alpha = 0,
  family = "cox",
  lambda = 10^seq(3, -2, by = -0.1)
)

m_beta_ridge_strong <- coef(m_ridge_strong, s = "lambda.min")
rmse_ridge_strong <- sqrt(mean((true_coefficients_strong - as.vector(m_beta_ridge_strong))^2))

m_cox_strong <- coxph(
  Surv(tstart, tstop, Delta) ~ X.1 + X.2 + X.3 + X.4 + X.5 + X.6 + X.7 + X.8 + X.9 + X.10,
  data = train_data_pseudo_strong
)

m_beta_cox_strong <- coef(m_cox_strong)
rmse_cox_n_500_strong <- sqrt(mean((true_coefficients_strong - as.vector(m_beta_cox_strong))^2))

# Generate n=2000 data with stronger signal
if (!file.exists(file_path_n_2000_strong)) {
  train_data_2000_strong <- generate_simulated_data(
    n = 2000,
    lambda_C = 0.1,
    eta_type = "10-dim-linear",
    CATE_type = "zero",
    seed_value = seed_value,
    linear_intercept = intercept_strong,
    linear_slope_multiplier = slope_multiplier_strong,
    verbose = 0
  )
  
  train_data_pseudo_2000_strong <- preprocess_data(
    single_data = train_data_2000_strong,
    run_time_varying = TRUE
  )
  
  saveRDS(
    list(
      train_data = train_data_2000_strong,
      train_data_pseudo = train_data_pseudo_2000_strong
    ),
    file = file_path_n_2000_strong
  )
} else {
  loaded_data_2000_strong <- readRDS(file_path_n_2000_strong)
  train_data_pseudo_2000_strong <- loaded_data_2000_strong$train_data_pseudo
}

m_cox_n_2000_strong <- coxph(
  Surv(tstart, tstop, Delta) ~ X.1 + X.2 + X.3 + X.4 + X.5 + X.6 + X.7 + X.8 + X.9 + X.10,
  data = train_data_pseudo_2000_strong
)

m_beta_cox_n_2000_strong <- coef(m_cox_n_2000_strong)
rmse_cox_n_2000_strong <- sqrt(mean((true_coefficients_strong - as.vector(m_beta_cox_n_2000_strong))^2))

print_coef_comparison(true_coefficients_strong, m_beta_default_strong, "Lasso Default Lambda (Strong Signal)")
print_coef_comparison(true_coefficients_strong, m_beta_custom_strong, "Lasso Custom Lambda (Strong Signal)")
print_coef_comparison(true_coefficients_strong, m_beta_ridge_strong, "Ridge Regression (Strong Signal)")
print_coef_comparison(true_coefficients_strong, m_beta_cox_strong, "Cox Regression (n=500, Strong Signal)")
print_coef_comparison(true_coefficients_strong, m_beta_cox_n_2000_strong, "Cox Regression (n=2000, Strong Signal)")
```

```{r combined-results-table}
results_df <- data.frame(
  Model = c("Lasso (Default λ)", "Lasso (Custom λ)", "Ridge", "Cox (n=500)", "Cox (n=2000)"),
  RMSE_Weak = c(rmse_lasso_default, rmse_lasso_custom, rmse_ridge, rmse_cox_n_500, rmse_cox_n_2000),
  RMSE_Strong = c(rmse_lasso_default_strong, rmse_lasso_custom_strong, rmse_ridge_strong, 
                 rmse_cox_n_500_strong, rmse_cox_n_2000_strong)
)

kable(results_df, 
      caption = "RMSE Comparison Across Models and Signal Strengths",
      col.names = c("Model", "RMSE (Weak Signal)", "RMSE (Strong Signal)"),
      digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Conclusions

1. Parameter strength significantly impacts model performance:
   - Weak signal (intercept = -1.5, slope = -0.5): Models struggle to recover true coefficients
   - Strong signal (intercept = 3, slope = 2.5): Models show markedly improved coefficient recovery

2. Sample size effects:
   - Larger sample size (n=2000) consistently improves coefficient estimation
   - Improvement is more pronounced with stronger signal parameters

3. Model comparison:
   - Ridge regression shows better stability across both parameter settings
   - Cox regression with n=2000 achieves the best performance with strong signals
   - Lasso tends to be more sensitive to signal strength

4. The strong parameter setting demonstrates that the DGP can effectively recover true coefficients when the signal-to-noise ratio is sufficiently high, validating the modeling approach.

5. Don't trust other people's regression coefficient that much. We can start there, but will likely tune ours.


## To do next
1. From the true vs estimated coefficient print out, we can see RMSE is not a good measure -- change it to a) differentce proportion to the true parameter; b) max component wise error 
-- This is done in function "calculate_min_relative_error" scripts/TV-CSL/tests/test-helper.R. To do is to implement that metric here.