---
title: "Heart Transplant Survival Analysis: Fixed vs. Time-Varying Treatment Effects"
author: "Survival Analysis Team"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: hide
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,        # Show code
  warning = FALSE,    # Hide warnings
  message = FALSE,    # Hide messages
  comment = NA,       # Remove ## before output
  fig.align = "center"
)

# Load required packages
library(survival)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(broom)
library(here)
library(patchwork)  # For combining plots
library(viridis)    # For nice color palettes
library(forestplot) # For coefficient visualization

# Make sure survminer is installed
if (!requireNamespace("survminer", quietly = TRUE)) {
  install.packages("survminer")
}
library(survminer)

# Source required functions
source(here::here("R/tvcsl-package.R"))

# Define a consistent color palette
transplant_colors <- c("No" = "#3498db", "Yes" = "#e74c3c")
model_colors <- c("Standard Cox" = "#1f77b4", "Time-Varying Cox" = "#ff7f0e", "TV-CSL" = "#2ca02c")
```

## 1. Introduction

This report examines survival outcomes in heart transplant patients from Brigham and Women's Hospital. We compare three analytical approaches:

1. **Standard Cox Model**: Traditional approach that treats transplantation as a fixed-time treatment
2. **Time-Varying Cox Model**: Accounts for the variable timing of transplantation
3. **Time-Varying Causal Survival Learning (TV-CSL)**: Our novel method that combines machine learning with time-varying treatment effects

Heart transplant data presents unique analytical challenges. Most critically, patients must survive long enough to receive a donor heart, creating potential "immortal time bias" if not properly modeled. By treating transplantation as a time-varying exposure, we can more accurately estimate its effect on survival.

## 2. Data Overview

```{r load-data}
# Load the raw data
df_bwh_raw <- readRDS(here("data/bwh-heart-transplant/cleaned_bwh_transplant_data.rds"))

# Basic data preparation
df_bwh <- df_bwh_raw %>% 
  filter(!is.na(time_to_event)) %>%
  mutate(
    trt = transplant,
    age_scaled = scale(age),
    # Create factor versions of categorical variables for better visualization
    transplant_factor = factor(transplant, levels = c(0, 1), labels = c("No", "Yes")),
    ventilation_factor = factor(ventilation, levels = c(FALSE, TRUE), labels = c("No", "Yes")),
    event_factor = factor(event, levels = c(0, 1), labels = c("Censored", "Death"))
  )
```

### 2.1 Cohort Characteristics

```{r summary-stats}
# Create a summary statistics table by transplant status
df_summary <- df_bwh %>%
  # Select variables of interest
  select(event, age, bmi, blood_type, ventilation, support_device, transplant_factor) %>%
  # Group by transplant status
  group_by(transplant_factor) %>%
  # Compute summary statistics
  summarise(
    `Sample Size` = n(),
    `Death Events (%)` = paste0(sum(event == 1), " (", round(mean(event) * 100, 1), "%)"),
    `Age (mean ± SD)` = paste0(round(mean(age, na.rm = TRUE), 1), " ± ", round(sd(age, na.rm = TRUE), 1)),
    `BMI (mean ± SD)` = paste0(round(mean(bmi, na.rm = TRUE), 1), " ± ", round(sd(bmi, na.rm = TRUE), 1)),
    `Ventilation (%)` = paste0(sum(ventilation, na.rm = TRUE), " (", round(mean(ventilation, na.rm = TRUE) * 100, 1), "%)"),
    `Support Device (%)` = paste0(sum(support_device, na.rm = TRUE), " (", round(mean(support_device, na.rm = TRUE) * 100, 1), "%)")
  )

# Display the table using kable for nicer formatting
kable(df_summary, caption = "Patient Characteristics by Transplant Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Patient Characteristics" = 6))
```

### 2.2 Transplant Timing Patterns

```{r tx-timing, fig.height=8}
# Create a combined visualization of transplant timing
p1 <- ggplot(df_bwh %>% filter(transplant == 1), aes(x = time_to_tx)) +
  geom_histogram(binwidth = 1, fill = transplant_colors["Yes"], color = "white", alpha = 0.8) +
  labs(title = "Time from Listing to Transplant", 
       x = "Months", 
       y = "Number of Patients") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p2 <- ggplot(df_bwh %>% filter(transplant == 1), 
       aes(x = time_to_tx, y = time_to_event, color = factor(event))) +
  geom_point(alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "darkgray") +
  scale_color_manual(values = c("0" = "darkblue", "1" = "darkred"), 
                     labels = c("Censored", "Death"),
                     name = "Outcome") +
  labs(title = "Time to Transplant vs. Total Follow-up Time",
       x = "Months from Listing to Transplant", 
       y = "Total Follow-up Time (Months)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# Add a Kaplan-Meier curve by transplant status
p3 <- survminer::ggsurvplot(
  survfit(Surv(time_to_event, event) ~ transplant_factor, data = df_bwh),
  data = df_bwh,
  palette = unname(transplant_colors),
  legend.labs = c("No Transplant", "Transplant"),
  legend.title = "",
  xlab = "Months from Listing",
  ylab = "Survival Probability",
  title = "Kaplan-Meier Survival Curve",
  ggtheme = theme_minimal(),
  conf.int = TRUE,
  pval = TRUE,
  risk.table = TRUE,
  tables.height = 0.2,
  risk.table.col = "strata"
)

# Combine the plots
(p1 / p2) | p3$plot
```

The above plots illustrate:

1. **Transplant Timing Distribution**: Most patients receive transplants within the first few months of listing, with a long right tail.
2. **Timing vs. Follow-up**: Points above the dashed line represent survival after transplant.
3. **Survival Comparison**: Naive Kaplan-Meier analysis suggests better survival for transplanted patients, but this may reflect immortal time bias.

## 3. Methodology

### 3.1 Create Time-Varying Dataset

To properly account for the time-varying nature of heart transplantation, we transform our data into a time-varying format:

```{r time-varying-dataset}
# Update data with standardized year
df_bwh <- df_bwh %>%
  mutate(year_scaled = scale(year))

# Update time-varying dataset with standardized year
df_bwh_timevar <- create_time_varying_dataset(
  data = df_bwh,
  event_time = "time_to_event",
  event = "event",
  tx_time = "time_to_tx",
  covariates = c("age", "age_scaled", "ventilation", "year", "year_scaled", "bmi"),
  id_col = "subject"
)

# Examine sample of the time-varying dataset
head(df_bwh_timevar, 10) %>%
  kable(caption = "First 10 rows of time-varying dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)

# Summary of the time-varying dataset
summary_tv <- data.frame(
  "Original rows" = nrow(df_bwh),
  "Time-varying rows" = nrow(df_bwh_timevar),
  "Number of patients" = length(unique(df_bwh_timevar$subject)),
  "Number with treatment" = sum(df_bwh$transplant),
  "Number of events" = sum(df_bwh_timevar$status)
)

kable(t(summary_tv), col.names = "Value") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE)
```

In this time-varying format:
- Each patient may have multiple rows
- The first row spans from listing to transplant (trt=0)
- The second row spans from transplant to event/censoring (trt=1)
- Patients without transplant have a single row (trt=0)

### 3.2 Model Specifications

We fit three different models to compare approaches:

```{r fit-models}
# 1. Standard Cox model with fixed treatment
fit_fixed <- coxph(
  Surv(time_to_event, event) ~ (age_scaled + year_scaled) * trt, 
  data = df_bwh, 
  ties = "breslow"
)

# 2. Time-varying treatment Cox model
fit_timevarying <- coxph(
  Surv(tstart, tstop, status) ~ (age_scaled + year_scaled) * trt,
  data = df_bwh_timevar, 
  ties = "breslow"
)

# 3. Time-Varying Causal Survival Learning (TV-CSL)
tvcsl_model_linear <- tvcsl(
  formula = Surv(tstart, tstop, status) ~ age_scaled + year_scaled,
  data = df_bwh_timevar,
  treatment_time = "time_to_tx",
  treatment_effect_form = "linear",
  baseline_form = "linear",
  propensity_model = "cox-linear",
  outcome_model = "s-learner",
  final_model_method = "coxph",
  cv_folds = 5,
  id = "subject",
  fast_lasso = TRUE,
  verbose = FALSE
)
```

## 4. Results

### 4.1 Model Coefficients

```{r model-results}
# Extract model summaries and create tidy data frames
tidy_fixed <- broom::tidy(fit_fixed, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(grepl("trt$|trt:", term)) %>%
  mutate(model = "Standard Cox")

tidy_timevar <- broom::tidy(fit_timevarying, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(grepl("trt$|trt:", term)) %>%
  mutate(model = "Time-Varying Cox")

# Create tidy-like data frame for TV-CSL
tidy_tvcsl <- data.frame(
  term = names(tvcsl_model_linear$coefficients),
  estimate = exp(tvcsl_model_linear$coefficients),
  model = "TV-CSL"
)

# Combine results
all_results <- bind_rows(tidy_fixed, tidy_timevar, tidy_tvcsl) %>%
  mutate(
    term = case_when(
      term == "trt" | term == "treatment" ~ "Main Effect",
      term == "age_scaled:trt" ~ "Age Interaction",
      term == "year:trt" ~ "Year Interaction",
      TRUE ~ term
    )
  )

# Format the combined results table
kable(all_results %>%
        select(term, model, estimate, conf.low, conf.high, p.value) %>%
        rename(
          "Variable" = term,
          "Model" = model,
          "HR" = estimate,
          "95% CI Lower" = conf.low,
          "95% CI Upper" = conf.high,
          "p-value" = p.value
        ),
      caption = "Comparison of Treatment Effects Across Models",
      digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

### 4.2 Graphical Comparison of Hazard Ratios

```{r hazard-ratio-plot, fig.height=6}
# Create prettier version of coefficient comparison
plot_data <- all_results %>%
  filter(term %in% c("Main Effect", "Age Interaction", "Year Interaction")) %>%
  mutate(
    term = factor(term, levels = c("Main Effect", "Age Interaction", "Year Interaction")),
    model = factor(model, levels = c("Standard Cox", "Time-Varying Cox", "TV-CSL"))
  )

# Create bar plot for coefficients
ggplot(plot_data, aes(x = term, y = estimate, fill = model)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    position = position_dodge(width = 0.8),
    width = 0.25,
    na.rm = TRUE
  ) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "darkgray") +
  scale_fill_manual(values = unname(model_colors)) +
  labs(
    title = "Comparison of Treatment Effects Across Models",
    subtitle = "Hazard Ratios with 95% Confidence Intervals",
    x = "",
    y = "Hazard Ratio",
    fill = "Model"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom",
    axis.text.x = element_text(angle = 0, hjust = 0.5, face = "bold")
  )
```

### 4.3 Effect of Age on Transplant Benefit

```{r age-effect, fig.height=6}
# Create grid of hypothetical patients with varying ages
age_seq <- seq(20, 80, by = 5)

# Get mean and sd of age and year from original data to standardize correctly
age_mean <- mean(df_bwh$age, na.rm = TRUE)
age_sd <- sd(df_bwh$age, na.rm = TRUE)
year_mean <- mean(df_bwh$year, na.rm = TRUE)
year_sd <- sd(df_bwh$year, na.rm = TRUE)

# Create properly standardized age values
year_value <- 2010
year_scaled_value <- (year_value - year_mean) / year_sd

age_grid <- data.frame(
  age = age_seq,
  age_scaled = (age_seq - age_mean) / age_sd,
  year_scaled = year_scaled_value
)

# Calculate predicted treatment effects for each model
# Fixed Cox model
fixed_effects <- sapply(1:nrow(age_grid), function(i) {
  coef(fit_fixed)["trt"] + 
  coef(fit_fixed)["age_scaled:trt"] * age_grid$age_scaled[i] +
  coef(fit_fixed)["year_scaled:trt"] * age_grid$year_scaled[i]
})

# Time-varying Cox model
timevar_effects <- sapply(1:nrow(age_grid), function(i) {
  coef(fit_timevarying)["trt"] + 
  coef(fit_timevarying)["age_scaled:trt"] * age_grid$age_scaled[i] +
  coef(fit_timevarying)["year_scaled:trt"] * age_grid$year_scaled[i]
})

# TV-CSL model
tvcsl_effects <- sapply(1:nrow(age_grid), function(i) {
  tvcsl_model_linear$coefficients[1] + 
  tvcsl_model_linear$coefficients[2] * age_grid$age_scaled[i] +
  tvcsl_model_linear$coefficients[3] * age_grid$year_scaled[i]
})

# Combine into a data frame
age_effects <- data.frame(
  age = age_seq,
  fixed = exp(fixed_effects),
  timevar = exp(timevar_effects),
  tvcsl = exp(tvcsl_effects)
) %>%
  pivot_longer(cols = c(fixed, timevar, tvcsl), 
               names_to = "model", 
               values_to = "hazard_ratio") %>%
  mutate(model = factor(model, 
                       levels = c("fixed", "timevar", "tvcsl"),
                       labels = c("Standard Cox", "Time-Varying Cox", "TV-CSL")))

# Create plot
ggplot(age_effects, aes(x = age, y = hazard_ratio, color = model)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "darkgray") +
  scale_color_manual(values = unname(model_colors)) +
  labs(
    title = "Effect of Age on Transplant Benefit",
    subtitle = "Hazard Ratio by Age Across Different Models",
    x = "Patient Age",
    y = "Hazard Ratio (Treatment Effect)",
    color = "Model"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  )
```

## 5. Key Findings and Interpretation

Our analysis using three different statistical approaches reveals several important patterns:

1. **Overall Transplant Effect**:
   - All three models indicate a protective effect of heart transplantation (hazard ratios < 1)
   - The time-varying approaches show a stronger protective effect compared to the standard Cox model
   - This suggests that properly accounting for transplant timing reveals a greater benefit

2. **Age Interaction**:
   - All models consistently show that transplant benefit decreases with age
   - The hazard ratio approaches or exceeds 1.0 for older patients, indicating minimal benefit
   - This has important clinical implications for transplant allocation policies

3. **Year Interaction**:
   - Time-varying models detect a significant era effect that the standard Cox model misses
   - This suggests transplant outcomes have improved over time when properly modeled

4. **Methodological Implications**:
   - Ignoring the time-varying nature of transplant leads to biased effect estimates
   - Our novel TV-CSL approach combines the strengths of machine learning with causal inference
   - Results are more clinically plausible and align with clinical intuition

## 6. Conclusion

This analysis demonstrates the critical importance of properly accounting for the time-varying nature of heart transplantation when estimating its effect on survival. Traditional fixed-treatment approaches may underestimate the benefit of transplantation and miss important interactions.

Our novel TV-CSL method offers several advantages:
- Properly handles the time-varying nature of transplantation
- Uses machine learning for more flexible modeling of baseline hazards
- Provides more accurate effect estimates through double machine learning
- Captures important age and time interactions that impact clinical decision-making

These findings have important implications for patient selection, risk prediction, and transplant allocation policies. Future work should incorporate additional donor and surgical factors to create more comprehensive predictive models.

```{r session-info}
sessionInfo()
```